Grove: A Convergent Collaborative Structure-Editor Calculus

### 2021-08-19 grove-paper-3-formalism.tex
edits with Cyrus
fiill in details of definitions
start merging appendix defs into section 3

We will now make the intuitions developed in the previous section precise
by defining a collaborative structure editor calculus called Grove.
We begin in Sec.~\ref{sub:Syntax} with the syntax, then discuss
the internal graph representation in Sec.~\ref{sub:Convergent Graphs},
...

TODO: point of this section is commutativity theorem

TODO: (though as we show in \autoref{sec:Formalism} creating a vertex is implicit in edge creation)

The formalism is rather simple, \autoref{sec:Grove By Example} shows
that it is sufficient to handle many common situations,
and this simplicity aids in the predictability.

@@

The language we are modeling is shown in \autoref{fig:Base syntax}, standard for a simply-typed lambda calculus except that all sorts have empty holes (represented by $\hole$) and conflict holes (of the form $\conflictHole{q,\ldots,q}{v}{p}$).
In particular, the argument of a lambda can be either a bound variable or a hole.
Additionally, all non-hole forms have a superscripted identifier $u \in \U$.

% Turn that into a constructor set K + position set P + arity function (has
% type: K -> $\wp$(P)).

% TODO: replace positions with macros

\figureArity

@@

\subsubsection{Graphs}%
\label{sub:Graphs}

A graph $G : \E \rightarrow \Sigma$ is a function from edges to edge states,
where $\E = \U \times \V \times \P \times \V$,
unique IDs are drawn from some suitable set $\U$ equipped with a total ordering $\leq$,
vertices are drawn from $\V$,
positions are drawn from $\P$,
and edge states are drawn from $\Sigma$, all defined below.

Each edge $\e = (u, v, p, v^\prime)$ has a unique name, $u \in \U$, and connects
the child position $p$ of originating vertex $v$ to destination vertex
$v^\prime$. The state of each edge, $\e$, is determined by the corresponding
edge state, $G(\e) \in \Sigma = \{ \bot, \Plus, \Minus \}$. When $G(\e) = \bot$,
$\e$ has not been created or deleted. When $G(\e)$ is $\Plus$, the edge has been
created. When $G(\e)$ is $\Minus$, the edge has been deleted. The total
ordering $\bot \sqsubset \Plus \sqsubset \Minus$ forms a lattice over $\Sigma$.

A vertex $v = (u, k)$ identifies a distinct instance of some constructor $k \in
\K$, corresponding to a non-hole form defined in \autoref{fig:Base syntax}. Each
contructor has an arity, which we define as a set of positions. A position $p
\in \P$ identifies a location at some originating vertex from which edges to
destination vertices may originate. The $\arityOp$ function in
\autoref{fig:Constructors, Indexes and Arity} gives the complete mapping from
constructors to potential edge positions with their corresponding sorts. Every
graph contains a distinguished root vertex, root constructor, and root position,
respectively named $v_0$, $k_0$, and $p_0$.

% TODO: add a lead-in sentence here

@@

Trees are often easier to work with than arbitrary graphs directly, for example to present the graph to users or for type checking.
A graph can be decomposed into a collection of expressions governed by the grammar in \autoref{fig:Base syntax}, which we organize into a structure called a grove.

@@

\figureDecompositionDefHelpers

>>>>

We will now make the intuitions developed in the previous section precise by defining a collaborative structure editor calculus called Grove.
This section
  gives the syntax of our source language,
  discusses the internal representations of graphs and terms,
  presents a commutative graph edit semantics,
  and introduces the user action semantics.

% We begin in Sec.~\ref{sub:Syntax} with the syntax,
% then discuss the internal graph representation in Sec.~\ref{sub:Convergent Graphs},
% ...

% TODO: point of this section is commutativity theorem

% TODO: (though as we show in \autoref{sec:Formalism} creating a vertex is implicit in edge creation)

% The formalism is rather simple, \autoref{sec:Grove By Example} shows
% that it is sufficient to handle many common situations,
% and this simplicity aids in the predictability.

@@

Our source language (\autoref{fig:Syntax}) is a simply typed lambda calculus with references, empty holes, and conflict holes.
The syntax of terms is sorted into expressions, types, and patterns.
Expressions include
  variables $\eVar{G}{x}$,
  lambda abstractions with pattern arguments $\eFun{G}{q}{\tau}{e}$,
  applications $\eApp{G}{e_1}{e_2}$,
  and numbers $\eNum{G}{n}$ with addition $\ePlus{G}{e_1}{e_2}$ and multiplication $\eTimes{G}{e_1}{e_2}$.
Types include
  arrows $\tArrow{G}{\tau_1}{\tau_2}$
  and the type of numbers $\tNum{G}$.
Patterns include variables $\pVar{G}{x}$.
Each sort of term also contains
multiparent references $\multiVertex{\e}$,
unicycle references $\cycleVertex{\e}$,
conflict holes $\conflictHole{t_i}_{i < n}$,
and empty holes $\hole$.
Note that for conflict holes, it must be the case that $n \geq 2$.
We identify conflict holes up to reordering.

Each term that is not a conflict hole carries an annotation.
For terms that are not holes, the annotation corresponds to the graph or edge it recomposes to.
Multiparent references correspond to edges that target
  \emph{multiparent roots}---vertices targeted by multiple edges.
Unicycle references correspond to edges that target
  \emph{unicycle roots}---vertices that are their own least ancestor.
Empty holes, which denote the absence of any corresponding structure,
are annotated with the source needed to construct terms in their place.

Conflict holes correspond to multiple edges with a common source.
Since conflict holes cannot be nested or contain empty holes,
each term in a conflict hole is annotated with the graph or edge it recomposes to,
and the combined annotations correspond to the graph the entire conflict hole recomposes to.

% \figureArity

@@

Let $\U$ be a set of \emph{unique identifiers} equipped with a total ordering $\leq$,
and $\K$ the \emph{constructors} of terms in the source language,
and $\P$ the \emph{positions} of subterms in the source language,
and $\Sigma = \SetOf{\bot, \Plus, \Minus}$ the
  \emph{edge states} equipped with a total ordering $\bot \sqsubset \Plus \sqsubset \Minus$
  that forms a lattice over $\Sigma$.
Now, let $\V = \U \times \K$ all possible \emph{vertices},
and $\S = \V \times \P$ all possible \emph{sources},
and $\E = \U \times \S \times \V$ all possible \emph{edges}.

A \emph{graph} $G : \E \rightarrow \Sigma$ is a function from edges to edge states.
Graphs that represent a complete edit state have a distinguished
  root vertex $\rootVertex$
  and root position $\rootPosition$.
The \emph{inedges} of $G$ are the edges on which $G$ returns $\Plus$.
The \emph{outedges} of a source are the inedges that have the source.

An \emph{edge} $\e{=}(u, (v, p), v^\prime) \in \E$ represents a directed multi-edge
identified by $u \in \U$, originating from source $(v, p) \in \S$, with destination vertex $v' \in \V$.

A vertex $v = (u, k) \in \V$ is an instance of constructor $k \in \K$ identified by $u \in \U$.
Vertices correspond to \emph{constructable terms}---terms that are not references or holes.
The \emph{inedges} of $v$ are the inedges that target $v$.
The \emph{ingraph} of $v$ is the subgraph on inedges that target $v$.
We write $G_v$ to denote the ingraph of $v$.
The \emph{parents} of $v$ are the source vertices of the inedges of $v$.
The \emph{ancestors} of $v$ is defined inductively as the parent vertices of $v$ plus their ancestors.
The \emph{least ancestor} of $v$ is the ancestor with the smallest unique identifier.

% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %

\subsubsection{Well sorted graphs}
\label{sub:Well sorted graphs}

% TODO: where does this go?
% 
% Each contructor has an arity, which we define as a set of position-sort pairs. A position $p
% \in \P$ identifies a location at some originating vertex from which edges to
% destination vertices may originate. The $\arityOp$ function in
% \autoref{fig:Constructors, Indexes and Arity} gives the complete mapping from
% constructors to potential edge positions with their corresponding sorts.

@@

Trees are often easier to work with than arbitrary graphs directly, for example to present the graph to users or for type checking.
A graph can be decomposed into a collection of expressions governed by the grammar in \autoref{fig:Syntax}, which we organize into a structure called a grove.

@@
